### 2014-10-06 Studies/2011_San_Francisco/Report/San_Francisco_Fire.tex
San Francisco: Edits to Sections 5 and 6

The results of the FDS simulations are discussed in the following two sections. The first section addresses the results of the simulation as they relate to the flow path and the conditions in the interior stairwell. The second section examines the model results and observed outcomes of the fire incident as they relate to tactics discussed in recent experimental research.

@@

The NIOSH report~\cite{NIOSH:Bowyer2} indicated that the interior stairwell on the front side of the structure was used to determine that the fire was located in the basement. From the simulation results (Section~\ref{sec:temperature}), the conditions in the interior stairwell were initially tenable. After the rear basement windows failed, the simulation results indicate a high-hazard area that exceeded the conditions of a Class~III exposure (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F) in the stairwell near the laundry room landing area.

After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure.

The arrows shown in Fig.~\ref{fig:flow_path_top} indicate the simulated flow direction of gases from the basement after the window failures. In this figure, the red arrows indicate elevated gas temperatures, and the blue arrows indicate relatively cooler gas temperatures. The arrow that transitions from red to blue indicates that the heated gases were cooling as they flowed away from the fire towards the front of the structure. The two arrows located at the door on the second floor (towards the front of the structure) indicate bi-directional flow. At that doorway, cool ambient air flowed towards the fire near the bottom of the door, and hot gases flowed away from the fire near the top of the door.

\begin{figure}[!ht]
\includegraphics[width=6.5in]{../Figures/Flow_Path_Top_Vel_462_s_First_Floor}
\includegraphics[width=6.5in]{../Figures/Flow_Path_Top_Vel_462_s_Basement_Floor}
\caption[Top view of the simulated flow path in the interior stairwell.]
{Top view of the simulated flow path on the first floor and basement floor 120~s after the first rear basement window failure, 1.2~m above the floor (mid-level height). The arrows indicate the direction of flow. Two arrows located at a door or window indicate bi-directional flow at a vent opening. The dashed arrow indicates a continuation of flow from the basement towards the first floor. The velocity contours use the same scale as Figs.~\ref{fig:smv_velocity} and \ref{fig:smv_velocity_vectors} with a range of 0~m/s to 9~m/s (0~mph to 20~mph).}
\label{fig:flow_path_top}
\end{figure}

@@

A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him/her at risk. Exposure of equipment to temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) represents a Class~III exposure~\cite{Donnelly2006}. Firefighters are at increased risk levels when encountering Class~III exposure conditions for more than 5~minutes~\cite{Donnelly2006}.

Figure~\ref{fig:smv_flow_path_top_1} shows a top view of the temperature contours (at a height of 1.2~m above the floor) on the first floor at a time of 2~s before the first rear basement window failed. At this time, the temperature of the gases in the interior stairwell exceeded 200~$^{\circ}$C (400~$^{\circ}$F).

@@

Figure~\ref{fig:smv_flow_path_top_2} shows a top view of the temperature contours (at a height of 1.2~m above the floor) on the first floor at a time of 60~s and 120~s after the first rear basement window failed. At a time of 60~s after the first rear basement window failed, the temperature of the gases in the interior stairwell exceeded 540~$^{\circ}$C (1000~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F). As more rear basement windows failed and provided ambient air to the growing basement fire, the temperatures in the interior stairwell continued to rise. At a time of 120~s after the first rear basement window failed, the temperature of the gases in the interior stairwell exceeded 760~$^{\circ}$C (1400~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

@@

Based on these results and the simulation results shown in Sections~\ref{sec:velocity} and \ref{sec:temperature}, after the rear basement windows failed, the simulated flow path temperatures in the interior stairwell were in excess of 760~$^{\circ}$C (1400~$^{\circ}$F) and the flow velocities were approximately 9~m/s (20 mph). The conditions in the interior stairwell changed from tenable to high-hazard very rapidly following the rear basement window failure. Within the structure, the exhaust gases in the flow path were moving towards the doors located on the front side of the structure via the interior stairwell. These simulated temperatures are consistent with the post-incident conditions that were documented in the interior stairwell.

@@

After the primary search was conducted, closing the interior doors (i.e., door control) could have inhibited the establishment of a flow path in the interior stairwell. A 360$^\circ$ scene size-up by the arriving firefighters could have helped determine the location of the fire in the basement and identify potential flow paths within the structure. The interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. Firefighters should avoid placing themselves within a flow path, where elevated temperatures and flow velocities can present hazardous conditions and increase the rate of heat transfer to firefighting gear via convection.

The timing of the interior attack in the basement occurred after delays in forcible entry operations on the exterior basement door. Suppression operations should be coordinated with interior operations to reduce thermal hazards related to flow paths within a structure. Ongoing research by NIST, Underwriters Laboratories (UL), and others has demonstrates that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure~\cite{madrzykowski2009fire, kerber2009fire}.

@@

Based on a review of these incidents, it is clear that fires with rapidly developing or changing flow paths are a significant hazard to the fire service. The development of (or changes to) a flow path could be caused by the failure of a component of the structure, such as a door, window, or portion of a ceiling, wall or floor. Environmental conditions such as wind can generate hazardous thermal conditions within a flow path. Uncoordinated ventilation can also be the cause of increased thermal hazards within a flow path.

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 760~$^{\circ}$C (1400~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

@@

The authors would like to thank XXXX. The authors would also like to thank John Culbertson of Central Valley Fire District, MT, for his assistance in compiling Table~\ref{tab:LODD}. Finally, the authors would like to thank Joseph Willi of the University of Illinois Urbana-Champaign for his assistance in creating the drawings of the structure.

>>>>

The results of the simulations are discussed in the following sections. Section~\ref{sec:simulated_flow_path} addresses the results of the simulation as they relate to the flow path that was established in the interior stairwell. Section~\ref{sec:assessing_the_hazard} examines the model results as they relate to the hazardous conditions and exposure temperatures in the interior stairwell. Section~\ref{sec:tactical_condsiderations} discusses tactical considerations and outcomes of the fire incident as they relate to the fire dynamics and flow path conditions that have been observed in recent experimental research.

@@

The NIOSH report~\cite{NIOSH:Bowyer2} indicated that the interior stairwell on the front side of the structure was initially used by firefighters to determine that the fire was located in the basement. From the simulation results (Section~\ref{sec:temperature}), the conditions in the interior stairwell were initially tenable. After the rear basement windows failed, the simulation results indicate a high-hazard area in the stairwell near the laundry room landing area that exceeded the conditions of a Class~III exposure (temperatures greater than 260~$^{\circ}$C or 500~$^{\circ}$F).

After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the interior stairwell. The two E26 firefighters were located in the flow path between the basement and the doors on the front side of the structure.

The arrows shown in Fig.~\ref{fig:flow_path_top} indicate the simulated flow direction of gases on the basement floor and first floor at mid-level height, 120~s after the first rear basement window failed. In this figure, the dashed arrow indicates a continuation of flow from the basement towards the first floor. The two arrows located at the door and windows on the basement floor (towards the rear of the structure) indicate bi-directional flow at a vent opening. At the basement door and windows, cool ambient air flowed towards the fire near the bottom of the openings, and hot gases and smoke flowed away from the fire near the top of the openings.

\begin{figure}[!ht]
\includegraphics[width=6.5in]{../Figures/Flow_Path_Top_Vel_462_s_First_Floor}
\includegraphics[width=6.5in]{../Figures/Flow_Path_Top_Vel_462_s_Basement_Floor}
\caption[Top view of the simulated flow path in the interior stairwell.]
{Top view of the simulated flow path on the first floor and basement floor at mid-level height, 120~s after the first rear basement window failure. The arrows indicate the direction of flow. The dashed arrow indicates a continuation of flow from the basement towards the first floor. Two arrows located at a door or window indicate bi-directional flow at a vent opening. The velocity contours use the same scale as Figs.~\ref{fig:smv_velocity} and \ref{fig:smv_velocity_vectors} with a range of 0~m/s to 9~m/s (0~mph to 20~mph).}
\label{fig:flow_path_top}
\end{figure}

@@

A person is susceptible to second-degree burn injuries if exposed to temperatures greater than 55~$^{\circ}$C (130~$^{\circ}$F)~\cite{contactburn}. Although firefighters wear protective gear, gear only offers a finite amount of protection. The polycarbonate in a self contained breathing apparatus begins to soften when the material temperature reaches approximately 140~$^{\circ}$C (284~$^{\circ}$F)~\cite{mensch2011emergency}. Structural fire fighting coats and pants are tested to withstand temperatures of 260~$^{\circ}$C (500~$^{\circ}$F)~\cite{nfpa2013standard}. Prolonged exposure to elevated temperatures can result in a significant amount of heat transferred to the firefighter, putting him or her at risk. Exposure of equipment to temperatures of 260~$^{\circ}$C (500~$^{\circ}$F) represents a Class~III exposure~\cite{Donnelly2006}. Firefighters are at increased risk levels when encountering Class~III exposure conditions for more than 5~minutes~\cite{Donnelly2006}.

Figure~\ref{fig:smv_flow_path_top_1} shows a top view of the temperature contours (at a height of 1.2~m above the floor) on the first floor, 2~s before the first rear basement window failed. At this time, the temperature of the gases in the interior stairwell exceeded 200~$^{\circ}$C (400~$^{\circ}$F).

@@

Figure~\ref{fig:smv_flow_path_top_2} shows a top view of the temperature contours (at a height of 1.2~m above the floor) on the first floor, 60~s and 120~s after the first rear basement window failed. At a time of 60~s after the first rear basement window failed, the temperature of the gases in the interior stairwell exceeded 540~$^{\circ}$C (1000~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F). As more rear basement windows failed and provided ambient air to the growing basement fire, the temperatures in the interior stairwell continued to rise. At a time of 120~s after the first rear basement window failed, the temperature of the gases in the interior stairwell exceeded 760~$^{\circ}$C (1400~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

@@

Based on these results and the simulation results shown in Sections~\ref{sec:velocity} and \ref{sec:temperature}, after the rear basement windows failed, the simulated flow path temperatures in the interior stairwell were in excess of 760~$^{\circ}$C (1400~$^{\circ}$F) and the flow velocities were approximately 9~m/s (20 mph). The conditions in the interior stairwell changed from tenable to high-hazard very rapidly following the rear basement window failure. Within the structure, the hot gases and smoke were moving  along the flow path from the basement towards the doors located on the front side of the structure via the interior stairwell. The simulated temperatures are consistent with the post-incident conditions that were documented in the interior stairwell.

@@

After the primary search was conducted, closing the interior doors (i.e., door control) could have inhibited the establishment of a flow path in the interior stairwell. A 360$^\circ$ scene size-up by the arriving firefighters could have helped determine the location of the fire in the basement and identify potential flow paths within the structure. The interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure and vent openings at the front of the structure. Firefighters should avoid placing themselves within a flow path where elevated temperatures and flow velocities can present hazardous conditions and increase the rate of heat transfer to firefighting gear via convection.

The timing of the interior attack in the basement occurred after delays in the forcible entry operations on the exterior basement door. Fire suppression efforts should be coordinated with interior operations and ventilation procedures to reduce thermal hazards related to flow paths within a structure. Ongoing research by NIST, Underwriters Laboratories (UL), and others has demonstrated that applying water from the exterior into the fire area of a structure (prior to the start of interior operations) can significantly improve the safety of firefighters by reducing the thermal hazard from the fire and reducing the potential for developing high velocity hot gas flows within the structure~\cite{madrzykowski2009fire, kerber2009fire}.

@@

Based on a review of these incidents, it is clear that fires with rapidly developing or changing flow paths are a significant hazard to the fire service. The development of (or changes to) a flow path could be caused by the failure of a component of the structure, such as a door, window, or portion of a ceiling, wall or floor. Environmental conditions such as wind can generate hazardous thermal conditions within a flow path. Uncoordinated ventilation procedures can also be the cause of increased thermal hazards within a flow path.

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows failed, a flow path was established between the basement living room area and the doors located on the front side of the structure (the front door and the garage door). The rear basement window failures resulted in a rapid change in the conditions within the flow path. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. In the interior stairwell, the flow velocities were approximately 9~m/s (20 mph) and the temperature of the gases was estimated to be in excess of 760~$^{\circ}$C (1400~$^{\circ}$F), which exceeds the Class~III exposure temperature of 260~$^{\circ}$C (500~$^{\circ}$F).

@@

The authors would like to thank XXXX. The authors would also like to thank John Culbertson of Central Valley Fire District, MT, for his assistance in compiling Table~\ref{tab:LODD}. Finally, the authors would like to thank Joseph Willi of the University of Illinois Urbana-Champaign for his assistance in creating dimensioned drawings of the structure.
